Automatic conversion of colloquial Finnish to standard Finnish

### 2015-01-25 finnormal-nodalida-2015/finnormal2015.tex
more stuff to 2.1
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@58576 72bbbca6-d526-0410-a7d9-f06f51895060

@@

Morphology: dialectal variation, shortening and merging together morphemes, e.g. \emph{en min\"{a}} `not I' $\rightarrow$ \emph{emm\"{a}} (`I don't').
Syntax: due to the spontaneous nature of colloquial language, syntactic structures have stuffs that are common to spoken language, such as ......

Sentence 1 is written in colloquial Finnish, and Sentence 2 is how it would be expressed in a standard way.

\ex
\begingl
\gla nelonenkin oli vaan silleen et fonotaksista p\"{a}\"{a}ttelin //
\glb fourth-{\sc also} was just like.that that phonotactics.{\sc ela} I.deduced //
% \glft `I deduced also the fourth from phonotactics' //
\endgl
\xe

\ex
\begingl
\gla p\"{a}\"{a}ttelin nelj\"{a}nnenkin fonotaksista //
\glb I.deduced fourth-{\sc also} phonotactics.{\sc ela} //
% \glft `I deduced also the fourth from phonotactics' //
\endgl
\xe

\cite{viinikka2013} describe the common meaning of the terms colloquial  (\emph{puhekieli}) and standard (\emph{yleiskieli} or \emph{kirjakieli}) Finnish: standard language follows the recommendations of kielenhuolto and is unified in morphology and vocabulary; colloquial language is everyday and free and shows local and idiolectal variation and doesn't need to follow the rules of kielenhuolto.
% ``Yleens\"{a} kirjakieli ymm\"{a}rret\"{a}\"{a}n samaksi kuin yleiskieli: kielenhuollon suositusten mukaiseksi standardoiduksi kieleksi, jonka sanasto on yleistajuista. Puhekieli puolestaan miellet\"{a}\"{a}n kirjakielt\"{a} arkisemmaksi ja tyylilt\"{a}\"{a}n vapaamuotoisemmaksi kielimuodoksi, jonka ei tarvitse noudattaa kirjakielen s\"{a}\"{a}ntöj\"{a} (esim. mie, h\"{a}nen koira, me kirjotetaan vs. min\"{a}, h\"{a}nen koiransa, me kirjoitamme).''

>>>>

\cite{viinikka2013} describe the common meaning of the terms colloquial (\emph{puhekieli}) and standard (\emph{yleiskieli} or \emph{kirjakieli}) Finnish: standard language follows the recommendations of kielenhuolto and is unified in morphology and vocabulary; colloquial language shows local and idiolectal variation, and has structures that are characteristic to spoken variety,  such as discourse particles and incomplete clauses.
Prototypically, colloquial language is spoken, everyday and spontaneous, and standard is written, planned and structured.

@@

Morphology: dialectal variation, shortening and merging of morphemes, e.g. \emph{en min\"{a}} `not I' $\rightarrow$ \emph{emm\"{a}} (`I don't').

Syntax: due to the spontaneous nature of colloquial language, syntactic structures are common for spoken language. The following example is from our data set, first sentence is the original colloquial version. The gloss shows the actual word-by-word translation, and the translation shows a similar style and register in English.

\ex
\begingl
\gla seiskakin oli vaan silleen et fonotaksista p\"{a}\"{a}ttelin //
\glb seven-{\sc also} was just like.that that phonotactics.{\sc ela} I.deduced //
\glft `also the seventh, it was like, I just deduced it from phonotactics' //
\endgl
\xe

Sentence 2 is translated in many ways. On the lexical level, the word \emph{seiska} is a colloquial name for `number 7', and in a literary standard it is replaced by the ordinal \emph{seitsemäs} `seventh'. The syntactic structure of the sentence is markedly spoken; the word \emph{seiska} is topicalised, and the main information ``deduced from phonotactics'' is a subordinate clause. 
The translation below is shorter and more precise, leaving just the main information.

\ex
\begingl
\gla p\"{a}\"{a}ttelin seitsem\"{a}nnenkin fonotaksista //
\glb I.deduced seventh-{\sc also} phonotactics.{\sc ela} //
\glft `I deduced also the seventh from phonotactics' //
\endgl
\xe

For the partial version, we didn't touch the syntactic structure, but changed morphology and vocabulary. Below is a version that shows the original and the partial.

\ex
\begingl
\gla seiskakin oli vaan silleen et fonotaksista p\"{a}\"{a}ttelin //
\glb seitsemäskin oli vain sill\"{a}\#lailla ett\"{a} fonotaksista p\"{a}\"{a}ttelin //
%\glb seventh-{\sc also} was just like.that that phonotactics.{\sc ela} I.deduced //
% \glft `also the seventh, it was like, I just deduced it from phonotactics' //
\endgl
\xe

% ``Yleens\"{a} kirjakieli ymm\"{a}rret\"{a}\"{a}n samaksi kuin yleiskieli: kielenhuollon suositusten mukaiseksi standardoiduksi kieleksi, jonka sanasto on yleistajuista. Puhekieli puolestaan miellet\"{a}\"{a}n kirjakielt\"{a} arkisemmaksi ja tyylilt\"{a}\"{a}n vapaamuotoisemmaksi kielimuodoksi, jonka ei tarvitse noudattaa kirjakielen s\"{a}\"{a}ntöj\"{a} (esim. mie, h\"{a}nen koira, me kirjotetaan vs. min\"{a}, h\"{a}nen koiransa, me kirjoitamme).''
